\propertitle

### 2019-07-22 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-22T11:30:14+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The context we consider is as follows. During an experimental session we
have recorded the spiking activities of $n$ neurons for a certain amount of
time. These neurons are our \enquote{sample} or \enquote{subnetwork}. Their
spikes are binned into $T$ time bins and binarized to $\set{0,1}$ values in
each bin. Call $\ya_{t}$ the number of neurons that fire during time bin
$t$, divided by the total number of neurons $n$; this is the
\emph{normalized total activity} of the sample, or just \enquote{activity}
for short. It is also the network-averaged activity of the neurons.
Obviously $\ya_{t} \in\set{0,1/n,\dotsc,(n-1)/n,1}$; if $\ya_{t}=0$, no
neuron spikes during bin $t$; if $\ya_{t}=1$, all spike at some point
during bin $t$. For brevity, let's say \enquote{at $t$} for \enquote{during
  time bin $t$}. From the activities $\set{\ya_{t}}$ we can count how often
the activity levels $\ya=0$, $\ya=1/n$, and so on appeared during the
recording, obtaining the distribution of measured relative frequencies
$(\yff_{\ya}) \defs \yf$. We can also consider the sample activity at time
bins \emph{outside} of the recorded range. Such activity is unknown to us,
of course.

\mynote{maybe move this paragraph to the intro?}Present-day technologies
enable the recording of neuronal activity from small brain areas \mynote{be
  more specific}. For many animal species, the neurons that are recorded
within the area are not -- and at present cannot be -- specifically chosen
from among the rest, owing to several limiting factors; for example,
limitations in how precisely electrodes are inserted or neurons are
targeted by viruses. In fact, the set of recorded neurons may even change
during very long recordings or across experimental sessions. We assume that
there's an area, comprising a network of $N$ neurons, for which we believe
that any other sample of size $n$ could have equally plausibly been
recorded instead of the sample of $n$ neurons that was actually recorded.
This is our \enquote{full network}.\mynote{how about calling it
  \enquote{the pool}?} The normalized total activity of these $N$ neurons
at $t$ is $\yA_{t}$. The relative frequencies of the various activity
levels during the recording were $(\yFF_{\yA}) \defs \yF$. We don't know
the values $\yA_{t}$ at each $t$, or the frequency distribution $\yF$. We
only know for certain that $\yA_{t} \in \set{0,1/N,\dotsc,(N-1)/N,1}$ and
that $N\yA_{t} \ge n\ya_{t}$ for obvious reasons. For the time being we
assume that we know $N$; in \sect*** we discuss the consequences of our
lack of precise knowledge about this number.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\pxx(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $na=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
provide the same information as the first $m$ power moments and vice versa:
they linearly determine each other because $\tbinom{n\ya}{m}$ is a
polynomial in $\ya$ of degree $m$. Fixing the ones is therefore equivalent
to fixing the others. But the normalized factorial moments have this
extremely convenient property: \emph{the first $n$ normalized factorial
  moments for a sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya)
=    \sum_{\yA} %m!\,
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pxx(\yA),
  % \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\pxx(\yA)$ is our distribution of belief about the full-network
activity.

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network by constraining its $m$th factorial moment to be equal
to the sample's recorded average of $\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$,
for as many $m$ as we please with $1\le m \le n$. In formulae these
constraints on $\px(\yA)$ are
\begin{equation}
  \label{eq:final_constraint}
\underbracket{\frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  \equiv
   \sum_{\ya}
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1} \yff_{\ya}}_{\text{measured moments}}
  =
   \underbracket{\sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \px(\yA)}_{\text{distribution moments}}
\end{equation}
for all $m$ we wish.

@@

The crucial point is formula~\eqref{eq:equal_m_factorialmoments}, on which
our method is based, and which asserts the equality of the factorial
moments of the full network and its sample, (or, equivalently, asserting
that the power moments for the one are linear functions of the power
moments for the other). This formula is only valid if our beliefs about the
activity of the sample  $\pxx(\ya)$ and that of the full network
$\pxx(\yA)$ are related by a hypergeometric distribution:
\begin{equation}
  \label{eq:hypergeometric_relation}
    \pxx(\ya)=
  \binom{n}{n\ya}\;\binom{\yNv-n}{\yNv \yA-n\ya}\;\binom{\yNv}{\yNv \yA}^{-1}
\;\pxx(\yA).
\end{equation}
This is a relation of \enquote{drawing without replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
In other words, $N$ must be such that we equally believe other samples of
size $n$ could have been recorded by the electrodes or the probe used. This
is indeed what we stated in \sect~\ref{sec:model}.

@@

@@

The question, then, is \emph{which features of the maximum-entropy
  frequency distribution are typical of the majority of plausible frequency
  distributions?} We can only answer for sure by using the full-fledged
probability calculus. A more complete study (in preparation) with the first
data set reveals that most of the plausible frequency distributions have
these important features in common with the maximum-entropy one:
\begin{itemize}
\item all activities $\yA \lesssim 0.05\%$ have practically zero frequencies;
\item there are two regions of activity levels with high frequencies,
  roughly separated by a trough of lower frequencies. The frequencies of
  the region on the left ($\yA \lesssim 0.018\%$) are higher than those of
  the region on the right ($\yA \gtrsim 0.018\%$).
\end{itemize}
But there are also differences. For example, plausible frequency
distributions may have three or four modes instead of just two; and these
modes are higher than of the the maximum-entropy distribution.

The second point is that our degrees of belief about the frequency
distribution for the full network depend not only on the measured data in
the sample, but also on our pre-data beliefs about the distribution. Which
assumptions lie behind the maximum-entropy approximation?

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-07-22T12:48:21+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The context we consider is as follows. During an experimental session we
have recorded the spiking activities of $n$ neurons for a certain amount of
time. These neurons are our \enquote{sample} or \enquote{subnetwork}. Their
spikes are binned into $T$ time bins and binarized to $\set{0,1}$ values in
each bin. Call $\ya_{t}$ the number of neurons that fire during time bin
$t$: this is the \emph{total activity} of the sample, or just
\enquote{activity} for short. Obviously $\ya_{t} \in\set{0,1,\dotsc,n}$; if
$\ya_{t}=0$, no neuron spikes during bin $t$; if $\ya_{t}=n$, all spike at
some point during bin $t$. For brevity, let's say \enquote{at $t$} for
\enquote{during time bin $t$}. If we divide the total activity by the
population size, $\ya/n \in \set{0,1/n,\dotsc,1}$ we have the normalized
total activity or population-averaged activity. From the activities
$\set{\ya_{t}}$ we can count how often the activity levels $\ya=0$,
$\ya=1$, and so on appeared during the recording, obtaining the
distribution of measured relative frequencies $(\yff_{\ya}) \defs \yf$. We
can also consider the sample activity at time bins \emph{outside} of the
recorded range. Such activity is unknown to us, of course.

\mynote{maybe move this paragraph to the intro?}Present-day technologies
enable the recording of neuronal activity from small brain areas \mynote{be
  more specific}. For many animal species, the neurons that are recorded
within the area are not -- and at present cannot be -- specifically chosen
from among the rest, owing to several limiting factors; for example,
limitations in how precisely electrodes are inserted or neurons are
targeted by viruses. In fact, the set of recorded neurons may even change
during very long recordings or across experimental sessions. We assume that
there's an area, comprising a network of $N$ neurons, for which we believe
that any other sample of size $n$ could have equally plausibly been
recorded instead of the sample of $n$ neurons that was actually recorded.
This is our \enquote{full network}.\mynote{how about calling it
  \enquote{the pool}?} The total activity of these $N$ neurons at $t$ is
$\yA_{t}$. The relative frequencies of the various activity levels during
the recording were $(\yFF_{\yA}) \defs \yF$. We don't know the values
$\yA_{t}$ at each $t$, or the frequency distribution $\yF$. We only know
for certain that $\yA_{t} \in \set{0, 1, \dotsc, N}$ and that
$\yA_{t} \ge \ya_{t}$ for obvious reasons. For the time being we assume
that we know $N$; in \sect*** we discuss the consequences of our lack of
precise knowledge about this number.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments} \citep{potts1953}. The $m$th normalized
factorial moment of a distribution $\pxx(\ya)$ for the activity of the
sample neurons is defined as the average
\begin{equation}
  % \E\condt[\Bigg]{
  %   \binom{\ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  % \equiv
  \sum_{\ya} %m!\,
  \binom{\ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya),
  \qquad 1\le m \le n
\end{equation}
% where $\px(\ya \| \yH)$ is the probability for activity $\ya$ given
% assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
% new bin.
This moment can be interpreted as the expectation of the number of distinct
$m$-tuples of simultaneously spiking neurons (within a bin's time width),
normalized by the number of distinct $m$-tuples. For example, with $m=2$,
if $na=4$ neurons spike in a network of $n=5$, we have $\tbinom{4}{2}=6$
distinct pairs of simultaneously spiking neurons, and the total number of
distinct pairs is $\tbinom{5}{2}=10$. The normalized number of spiking
pairs is therefore $6/10$. Note that the first $m$ factorial moments
provide the same information as the first $m$ power moments and vice versa:
they linearly determine each other because $\tbinom{\ya}{m}$ is a
polynomial in $\ya$ of degree $m$. Fixing the ones is therefore equivalent
to fixing the others. But the normalized factorial moments have this
extremely convenient property: \emph{the first $n$ normalized factorial
  moments for a sample and for the full network are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
    \sum_{\ya} %m!\,
  \binom{\ya}{m}\,  \binom{n}{m}^{-1}\, \pxx(\ya)
=    \sum_{\yA} %m!\,
  \binom{\yA}{m}\,  \binom{N}{m}^{-1}\, \pxx(\yA),
  % \E\condt[\big]{\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  % \E\condt[\big]{\tbinom{\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad 1\le m \le n,
\end{equation}
where $\pxx(\yA)$ is our distribution of belief about the full-network
activity.

We can therefore apply the maximum-entropy method to obtain a distribution
for the full network by constraining its $m$th factorial moment to be equal
to the sample's recorded average of $\tbinom{\ya}{m}\;\tbinom{n}{m}^{-1}$,
for as many $m$ as we please with $1\le m \le n$. In formulae these
constraints on $\px(\yA)$ are
\begin{equation}
  \label{eq:final_constraint}
\underbracket{\frac{1}{T}\sum_{t}\binom{\ya_{t}}{m}\,\binom{n}{m}^{-1}
  \equiv
   \sum_{\ya}
  \binom{\ya}{m}\,  \binom{n}{m}^{-1} \yff_{\ya}}_{\text{measured moments}}
  =
   \underbracket{\sum_{\yA}
  \binom{\yA}{m}\,  \binom{N}{m}^{-1}\, \px(\yA)}_{\text{distribution moments}}
\end{equation}
for all $m$ we wish.

@@

The crucial point is formula~\eqref{eq:equal_m_factorialmoments}, on which
our method is based, and which asserts the equality of the factorial
moments of the full network and its sample, (or, equivalently, asserting
that the power moments for the one are linear functions of the power
moments for the other). This formula is only valid if our beliefs about the
activity of the sample  $\pxx(\ya)$ and that of the full network
$\pxx(\yA)$ are related by a hypergeometric distribution:
\begin{equation}
  \label{eq:hypergeometric_relation}
    \pxx(\ya)=
  \binom{n}{\ya}\;\binom{N-n}{\yA-\ya}\;\binom{N}{\yA}^{-1}
\;\pxx(\yA).
\end{equation}
This is a relation of \enquote{drawing without replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}.
In other words, $N$ must be such that we equally believe other samples of
size $n$ could have been recorded by the electrodes or the probe used. This
is indeed what we stated in \sect~\ref{sec:model}.

@@

\section{Limitations and assumptions}
\label{sec:assumptions}

@@

The question, then, is \emph{which features of the maximum-entropy
  frequency distribution are typical of the majority of plausible frequency
  distributions?} We can only answer for sure by using the full-fledged
probability calculus. A more complete study (in preparation) with the first
data set reveals that most of the plausible frequency distributions have
these important features in common with the maximum-entropy one:
\begin{itemize}
\item all activities $\yA/N \lesssim 5\%$ have practically zero frequencies;
\item there are two regions of activity levels with high frequencies,
  roughly separated by a trough of lower frequencies. The frequencies of
  the region on the left ($\yA/N \lesssim 1.8\%$) are higher than those of
  the region on the right ($\yA/N \gtrsim 1.8\%$).
\end{itemize}
But there are also differences. For example, plausible frequency
distributions may have three or four modes instead of just two; and these
modes are higher than of the the maximum-entropy distribution.

The second point is that our degrees of belief about the frequency
distribution for the full network depend not only on the measured data in
the sample, but also on our pre-data beliefs $\yI$ about the distribution. Which
assumptions lead to the maximum-entropy result? This distribution appear
when our initial belief about the possible frequency distributions $\yF$ is
quantified by an entropic prior:
\begin{equation}
  \label{eq:entropic_prior}
  \pf(\yF \| \yI) \propto  \exp[-\yL\;\ysh(\yF;\yR)]
  \approx \binom{L}{L\yFF_{0}, \dotsc, L\yFF_{N}}\;
  \prod_{\yA}{\yRR_{\yA}}^{L\yFF_{\yA}}
\end{equation}
where $\ysh(\yF;\yR)\defd \sum_{\yA}\yFF_{\yA}\ln(\yFF_{\yA}/\yRR_{\yA})$
is the relative entropy or discrimination information
\citep{kullback1987,jaynes1963,hobson1969,hobsonetal1973}, $\yR$ is a
reference distribution, and $\yL$ a positive parameter.
